---
title: "Simple parametric statistics"
output:
  html_document:
    css: ../extras.css
    theme: cerulean
    highlight: tango
---

```{r, include = FALSE}
library(dplyr)
library(ggplot2)
```

You should work through the exercises step-by-step, following the instructions carefully. At various points we will interrupt the flow of instructions with a question. Make a note of your answers so that you can complete the MOLE quiz for this week.

### Eagle owls and Norway rats

```{r, eval = FALSE, echo = FALSE}
ratskulls <- read.csv("../data_csv/RATSKULL.CSV")
ggplot(ratskulls, aes(x = Weight)) + 
  geom_dotplot() + facet_wrap(~Season, ncol = 1)
t.test(Weight ~ Season, data = ratskulls)
```

A data set containing information about the sizes of Norway rat skulls in the pellets of Scandinavian eagle-owls is available in the RATSKULL.CSV file. The data comprise a column of rat skull sizes (measured in grams) and a column of codes indicating the season when a particular skull sample was taken. These data were collected in order to evaluate whether there is a difference between sizes of rats eaten in summer and winter, i.e. we want to know if there is a statistically significant difference between the mean rat skull sizes in the winter and summer samples.

Download the RATSKULL.CSV file from MOLE and place it in your working directory. Read the data in RATSKULL.CSV into an R data frame, remembering to assign it a name.

As always, start by *looking at the data* — both visually and in terms of its descriptive statistics:

**Inspection.** Use the `View` function and `dplyr` function `glimpse` to inspect the raw data. What are the names given to rat skull size variable and the season indicator variable? What values does the season indicator variable take?

**Descriptive statistics.** Use the appropriate `dplyr` functions (`group_by` and `summarise`) to calculate the sample size, sample mean and standard deviation of each sample. You will need to use the `mean`, `sd` and `n` functions to do this.

**Graphs.** Use `ggplot2` to construct a pair of dot plots, one above the other, to summarise the winter and summer skull size distributions. You will need to use `geom_dotplot` and `facet_wrap` to do this.

Using the dot plots, and the descriptive statistics, conduct an informal evaluation of the assumptions of the *t*-test. You should re-read the relevant section in the course book if you can't remember what these are.

```{block, type='do-something'}
**MOLE question**

Do you feel the data conform acceptably to the assumptions? If not, explain why.
```

Let's carry on, assuming that we are confident that it is OK to use a two sample *t*-test to compare the sample means. Use the R `t.test` function to carry out this evaluation now.

```{block, type='do-something'}
**MOLE question**

Write a concise but complete conclusion summarising the results of the test.

Is this what you expected from looking at the distributions of data in the two samples?

Suggest two possible biological reasons for the result you observe.
```


### Fungal infection in French beans

```{r, echo=FALSE}
sandy <- c(2.3, 2.4, 2.5, 2.6, 2.8, 2.7, 3.1, 2.3, 2.5)
clay  <- c(2.3, 2.5, 2.8, 3.2, 2.9, 3.1, 3.2)
```

```{r, echo=FALSE, eval=FALSE}
t.test(sandy, clay)
beans <- data.frame(glucosamine = c(sandy, clay),
                    soil = rep(c("sandy","clay"), c(length(sandy), length(clay))))
write.csv(beans, file = "../course-data/FRENCH_BEANS.CSV", row.names = FALSE)
```

A plant pathologist noticed that fungal infection in roots of French beans (*Phaseolus vulgaris*) was rather variable among crops and hypothesized that infection might be affected by the soil type: in particular whether the beans were grown on clayey or sandy soils. Root samples were taken from beans growing in each soil type and fungal infection was measured indirectly by measuring the amount of glucosamine in the roots. Glucosamine is a fungal sugar which is polymerised into chitin which forms the cell walls of most fungi.

The glucosamine concentrations ($\mu$g g$^{-1}$ root dry weight) recorded from the samples were:

| Sandy Soil | Clay Soil |
|:----------:|:---------:|
| 2.3 | 2.3 |
| 2.4 | 2.5 |
| 2.5 | 2.8 |
| 2.6 | 3.2 |
| 2.8 | 2.9 |
| 2.7 | 3.1 |
| 3.1 | 3.2 |
| 2.3 |     |
| 2.5 |     |

Enter these data into Excel, using a tidy data format. Remember to save it as a CSV file in your working directory. 

Read the data into an R data frame, inspect the data, generate some summary statistics (means, SDs and sample sizes) and then plot the data, just as you did in the last exercise. This should be quick to achieve if you copy and paste the code you produced above and edit this where required.

Use a *t*-test to determine whether there is a significant difference between the amount of infection of bean roots in the two different soils.

```{block, type='do-something'}
**MOLE question**

Make a note of the results:

Mean for plants on clay soil = ?

Mean for plants on sandy soil = ?

t = ?

d.f. = ? 

p = ? 
```

```{block, type='do-something'}
**MOLE question**

Write a statement of the result of the test suitable for inclusion in the results section of the plant pathologist's report.
```


### Plant and insect diversity

```{r, echo = FALSE, eval = FALSE}

set.seed(110189)
library(truncnorm)

x <- rtruncnorm(20, a = 0, b = Inf, mean = 4, sd = 1.5)#runif(20, 0, 6)
div.data <- data.frame(PlantDiv = x, InsectDiv = 0.6 * x + rnorm(20, 0, 1)) %>% 
  mutate(InsectDiv = ifelse(InsectDiv<0, InsectDiv + 1.2 * (0-InsectDiv),InsectDiv))

ggplot(div.data, aes(x=PlantDiv, y = InsectDiv)) + 
  geom_point()

write.csv(div.data, file = "../data_csv/DIVERSITY.CSV")

```

An ecologist was interested in whether there was an association between plant and herbivorous insect diversity in a moorland habitat. They assessed plant diversity within 20 quadrats. Pitfall traps placed within the quadrats were used to estimate herbivorous insect diversity.The DIVERSITY.CSV file contains data on plant diversity and herbivorous insect diversity. 

Explore the data using glimpse. Plot the data and decide whether you think that there is an association. 

```{block, type='do-something'}
**MOLE question**

Which test could you use to see if there was an association between plant and insect diversity?
```

Check whether the data meet the assumptions of the test. Carry out the appropriate test and look at the results.

```{block, type='do-something'}
**MOLE question**

Write a clear concise summary of your results.
```
