# Two-way ANCOVA in R

## Introduction

We'll use the sea otter predation example from the previous chapter to walk through how to carry out an ANCOVA in R.

```{block, type='do-something'}
**Walk through example**

You should begin working through the example from this point.
```

You need to download the OTTERS.CSV file from MOLE and place it in your working directory. Read the data into an R data frame, giving it the name `seaotters`. Make sure you have a look at the data before you proceed.

```{r, echo=FALSE}
seaotters <- read.csv(file = "./data_csv/OTTERS.CSV")
```

The data are laid out with the response variable in one column and two additional columns for the predictor variables: one contains the codes for the categorical variable, the other contains the values of the numeric variable. Thus, the first column (`Otters`) contains the sea otter abundances, the second column (`Location`) contains the codes for the study population (levels: 'Lagoon' and 'Bay'), and the third column (`Year`) contains the observation year (1993-2004).

## Visualising the data

As always we should start by visualising our data. 
```{r, echo=FALSE}
seaotters <- read.csv(file = "./data_csv/OTTERS.CSV")
```

```{r, fig.align='center', fig.width = 6, fig.height = 4}
ggplot(seaotters, aes(x = Year, y = Otters, colour = Location)) + 
  geom_point()
```

As we noted in the previous chapter the figure suggestes that sea otter abundances have declined in both locations, with a greater decline where sea otters were exposed to predation from killer whales (the bay location). 

We should also consider the assumptions of the ANCOVA. The scatter plot suggests that, within each location, the relationship between $x$ and $y$ is linear. The numeric predictor variable (study year) is measured on an interval scale and the response variable (otter abundance) is measured on ratio scale. Year is obviously measured without error. What about the independence assumption?

```{block, type='do-something'}
**Independence**

Can you think of any reasons why the independence assumption may be problematic in this example? Think about how the data have been collected---they are a time series of abundances in a pair of adjacent populations.
```

We'll assume the independence assumption has been met in these data. As with regression, the remaining assumptions are probably best measured using regression diagnostics after we've fitted the model. 


## Fitting an ANCOVA

As you should expect by this point, carrying out ANCOVA in R is a two step process. The first step is the model fitting step. This is where R calculates the best fit intercepts and slopes for each group (i.e. each location in this example), along with additional information needed to carry out the evaluation of significance in step two.

We carry out the model fitting step using the `lm` function:
```{r} 
otters.model <- lm(Otters ~ Location + Year + Location:Year, data = seaotters)
```
This is just more of the same old model fitting with `lm`. We assigned two arguments:

1. The first argument is a **formula**. The variable name on the left of the `~` must be the response variable (`Otters`) and the terms on the right must only include the two predictor variables (`Location` and `Year`).

2. The second argument is the name of the data frame that contains the variables listed in the formula (`seaotters`).

Let's unpack the formula we used:
```{r, eval = FALSE}
Otters ~ Location + Year + Location:Year
```
There are three terms, each separated by a `+` symbol: the two main effects (`Location` and `Year`) and their interaction (`Location:Year`). This tells R that we want to fit a model accounting for the main effects of study location and year, but that we also wish to include the interaction between these two variables. The `Location` term allows each line to cross the y-axis at a different point, the `Year` term allows the effect of year (the slope) to be non-zero, and the interaction term allows this slope to be different in each location. 

```{block, type='advanced-box'}
**How does R knows we want to carry out ANCOVA?**

Notice how similar fitting this ANCOVA was to fitting a two-way ANOVA. How does R know we want to use ANCOVA? You should be able to answer this question. R looks at what type of variables are on the right had side of the `~` in the formula. Since `Location` is a factor and `Year` is numeric, R fits an ANCOVA model. If both variables had been factors we fit a two-way ANOVA, and if both variables were numeric we would fit something called a multiple regression model. 
```

## Diagnostics

Before we go on to look at the *p*-values we should check the remaining assumptions using the diagnostics. We'll make the same plots as if we'd fitted a linear regression. First we'll evaluate the linearity assumption by constructing a **residuals vs. fitted values** plot. 

```{r, fig.align='center', fig.asp=1, fig.width=4}
plot(otters.model, add.smooth = FALSE, which = 1)
```

There's no evidence of a systematic trend here so the linearity assumption is fine. We'll move on to the normality assumption next, by making a **normal probability plot**. 

```{r, fig.align='center', fig.asp=1, fig.width=4}
plot(otters.model, which = 2)
```

This doesn't look great, very few of the points are on the dashed line and there appears to be a systematic trend away from the line. We'll carry on for now as we're just using this as an example of how to carry out an ANCOVA. If we were really interested in the results of this analysis we should consider transforming our response variable.

```{block, type='do-something'}
**Normality assumption**

Can you think of any reason that we might expect the residuals in these data not to be normally distributed? 
```

Finally, we'll consider the constant variance assumption using the **scale location plot**. 

```{r, fig.align='center', fig.asp=1, fig.width=4}
plot(otters.model, add.smooth = FALSE, which = 3)
```

Here, we're on the lookout for a systematic pattern in the size of the residuals and the fitted values---does the variability go up or down with the fitted values? There doesn't appear to be a strong pattern here.

## Interpreting the results

```{r, echo=FALSE}
anova.out <- capture.output(anova(otters.model))
```

Next, we use the `anova` function to determine whether the main effects and the interaction are significant, by passing it the name of the fitted regression model object (`otters.model`):
```{r}
anova(otters.model)
```
The first line reminds us that we are looking at an ANOVA table. Remember, this doesn't necessarily mean we are dealing with an ANOVA model---we are definitely examining an ANCOVA here. The second line reminds us what variable we analysed (i.e., the response variable). The critical part of the output is the table at the end:
```{r, echo = FALSE}
invisible(sapply(anova.out[4:8], function(line) cat(line, "\n")))
```
This summarises the parts of the analysis of variance calculations, as they apply to ANCOVA. These are: `Df` -- degrees of freedom, `Sum Sq` -- the sum of squares, `Mean Sq` -- the mean square, `F value` -- the *F*-statistic (i.e. variance ratio), `Pr(>F)` -- the p-value).

The *F*-statistics (variance ratios) are the key terms. When working with an ANCOVA, these relate to how much variability in the data is explained when we include each term in the model, taking into account the degrees of freedom it 'uses up'. Larger values indicate a stronger effect. The p-value gives the probability that the relationship could have arisen through sampling variation, if in fact there were no real association: a p-value of less than 0.05 indicates a less than 1 in 20 chance of the result being due to chance, and we take this as evidence that the relationship is real.

We need to interpret these *p*-values. The two main effects are significant (*p*<0.001), but the interaction is not (*p*=0.076). An ANOVA table tells us nothing about the direction of the effects---we have to plot the data to be able to do this. If we look back at the scatter plot, it is apparent that the significant main effects are supporting the observation that otter abundances are higher in the bay area, and that in general, otter abundances have declined over the course of the study. The interaction term is non-significant---though we only just missed the conventional *p*<0.05 cut off. We are forced to conclude that the data do not support the hypothesis that the population abundances in each location have declined by different amounts.

## Presenting the results 

We will need to provide a succinct factual summary of the analysis in the results section of the report:

> There were significant effects of location (ANCOVA: F=139.4, df=1,20, *p*<0.001) and year (F=20.4; df=1,20; p<0.001) on sea otter abundance. The interaction between location and year was not significant (F=3.5, df=1,20, p=0.076). Sea otter abundances were generally higher in Handae Bay, but declined by a similar amount in both locations during the study (Figure 1).

Notice that we never referred to 'treatments' in this summary. It does not make any sense to describe the variables in this data set as treatments, as we are describing the results from an observational study. Of course, there is nothing to stop us using ANCOVA to analyse experimental data if it is appropriate.

For presentation it is best to present the results as a figure. We can produce publication quality figure to summarise ANCOVA in much the same way as we summarise a fitted regression model. We are aiming to produce a figure that shows two pieces of information: a scatter plot and lines of best fit. We also want to differentiates the data and best fit lines for each location. We know how to produce a scatter plot, so the main challenge is to add the lines of best fit. We use the `predict` function to do this. 

To use `predict`, we have to let R know the set of values of the two predictor variables for which we want predictions (`Location` and `Year`). Sea otter abundances were evaluated from 1992 to 2003, so it makes sense to predict their abundances over this range. We also need to make distinct predictions for each location ('Lagoon' and 'Bay'). Therefore, the first step in making predictions is to generate a sequence of values for Year from 1992 to 2003 *for each location*, placing these alongside a location indicator inside a data frame. We have to use a new function, `expand.grid`, to do this:
```{r}
pred.data <- expand.grid(Year = 1992:2003, Location = c("Lagoon", "Bay"))
```
Remember that `1992:2003` produces a numeric vector in which the elements are a sequence of integers ('whole numbers') running from 1992 to 2003. 

That probably looks very cryptic at the moment. Look at `pred.data`:
```{r}
pred.data
```
The `expand.grid` produced a data frame with two variables called `Year` and `Location`. The rows of the data frame represent every pairwise combination of the two sets of values (i.e. the vectors) we passed to `expand.grid`. Notice that we gave the two arguments of `expand.grid` the exact same names as the predictor variables in the ANCOVA. **This is important**: the names in `pred.data` have to match the names of the predictor variables used in model, and every variable in the model has to be represented in `pred.data`. 

The next step is the same as the regression example. Once we have set up a data frame to predict from (`pred.data`) we are ready to use the `predict` function. We need to capture the predictions inside a data frame using `mutate`:
```{r}
pred.data <- mutate(pred.data, Otters = predict(otters.model, pred.data))
```

Look at the resulting data frame:
```{r, echo = TRUE}
pred.data
```
Notice that we gave the predictions the same name as the response variable in our ANCOVA model. This is convenient, because it means we don't have to set up any new aesthetic mappings when we use `ggplot2` in a moment. Notice that `pred.data` is set out just like the data frame containing the study data. It has three columns called `Otters`, `Location` and `Year`, but instead of raw data, it contains predictions from the model. 

Plotting these predictions along with the data is now very easy:
```{r, fig.align='center', fig.width = 6, fig.height = 4}
ggplot(pred.data, aes(x = Year, y = Otters, colour = Location)) + 
  geom_line() + geom_point(data = seaotters) + 
  xlab("Year") + ylab("Sea Otter Abundance")
```
Don't forget: we have to make `ggplot2` use the `seaotters` data (i.e. the raw data) when adding the points. 

Let's summarise what we did: 1) using `expand.grid`, we made a data frame with two columns containing the values of the predictor variables we want to make predictions at; 2) we then used `predict` to generate these predictions, adding them to the prediction data with `mutate`; 3) finally, we used `ggplot2` to plot the predicted values of the response variable against the numeric predictor variable, colouring the lines and points differently for each location.

This workflow is almost identical to that used to summarise a regression model---the only new trick here was the use of `expand.grid` to deal with the fact that now we have to manage two predictor variables.
