---
title: "One-way ANOVA"
output:
  html_document:
    css: ../extras.css
    theme: cerulean
    highlight: tango
---

Garden slugs of the same species and approximately the same size were given either a full dose, a half strength dose, or a one tenth strength dose of a molluscicide (‘Slugit’) dissolved in water. A control group were treated with just water. The rate of movement of the slugs on a moist wooden board after treatment was recorded (cm per min). There were 8 replicate slugs for each treatment.

The data are stored in a CSV file SLUGIT.CSV. The rate of movement is stored in the Distance variable and the treatment codes are in the SlugitTreat variable (levels: ‘full dose’, ‘half dose’, ‘one tenth dose’, ‘control’).

### First steps
As always start by loading the necessary packages.
```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(agricolae)
```

Set your working directory (using Session tab, Set working Directory, Choose Directory). This must be set to the folder containing the data set that you want to work with. Then read the data into R (remember to assign it a name).

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = "../data_csv")
```

```{r}
slug <- read.csv("SLUGIT.CSV")  
```

Start by having a quick look at the data. How many variables are there? How many observations? Is the focal variable (Distance) on a ratio or interval scale? 
```{r}
glimpse(slug)
```
Remember that for a one-way ANOVA the treatment (SlugitTreat) should not be stored as numbers (i.e. do not call the treatments "1, 2, 3,...") as this will result in a linear regression when the model is fitted below, rather than an ANOVA. 

The one-way ANOVA also requires that the experimental units are independent. This assumption should have been considered earlier, as part of the experimental design.

### Visualising the data

Next make a quick plot of the data (either a box and whiskers plot as below or a separate dot plot for each group). The two assumptions to check with this plot are that the within group variance is constant and that within each group the focal variable is normally distributed. 

```{r, fig.width=4, fig.asp=1.0, fig.align='center'}
ggplot(slug, aes(x = SlugitTreat, y = Distance)) + 
  geom_boxplot() 
```

Here, the variability within each group appears roughly similar. The distributions also look roughly symmetrical and there are no 'outliers' suggesting that the normal distribution assumption is not violated. You should come back to these assumptions again later, by carrying out the model diagnostics once the model has been fitted. 

### Fitting the model

Fit the model using the `lm` function, which takes a formula and a data frame containing the variables in the formula as its two arguments. Remember that in the formula the focal variable and the treatment go on the left and right of the ~ respectively.
```{r}
slug_model <- lm(Distance ~ SlugitTreat, data = slug)
```

### Check the model diagnostics

Remember to check the model diagnostics before looking at the statistical output.  

Check the normality assumption using a normal probability plot. Here, if the residuals are normally distributed they will be positioned on the dashed line of best fit. Note that we don't expect to see a perfect relationship, but the majority of the points should be very close to the line. 
```{r, fig.width=4, fig.asp=1.0, fig.align='center'}
plot(slug_model, which = 2)
```
This plot looks fine - the points don't deviate from the line in any systematic way.

Finally, check the constant variance assumption using a scale-location plot. Ideally there should be a flat relationship between the transformed residuals (y-axis) and the fitted values (x-axis). If the magnitude of the residuals either increases or decreases as the fitted values increase then the constant variance assumption is violated.
```{r, fig.width=4, fig.asp=1.0, fig.align='center'}
plot(slug_model, which = 3, add.smooth = FALSE)
```
There doesn't appear to be a systematic pattern in the residuals here so the constant variance assumption is met. 

Remember that if the data do not fit the assumptions of the one-way ANOVA the focal variable can be transformed. After fitting the model to the transformed data remember to check the diagnostics again to see if the assumptions are now met. If the assumptions are still not met it may be possible to use a non-parametric test instead. The non-parametric equivalent to a one-way ANOVA is the Kruskal-Wallis test. 

### Statistical output

Assuming the model assumptions were not violated now look at the model output. Use the `anova` function to calculate the F-ratio. This function takes the name of the fitted model as its only argument.

```{r}
anova(slug_model)
```

The table summarises the different parts of the *F*-test calculations: `Df` – degrees of freedom, `Sum Sq` – the sum of squares, `Mean Sq` – the mean square, `F value` – the *F*-statistic, `Pr(>F)` – the p-value. The *F*-statistic is the key term; larger values of the *F* statistic indicate clearer differences among the means of the different treatment groups. The p-value gives the probability that the observed differences between the means, or a more extreme difference, could have arisen through sampling variation under the null hypothesis. What is the null hypothesis: it is one of no effect of treatment, i.e. the null hypothesis is that all the means are the same. A p-value of less than 0.05 indicates the result is statistically significant (i.e. that there is a significant difference between the means of different treatment groups). Here, there does appear to be a significant effect of molluscicide dosage on slug activity as p<0.001.  

Also make a note of the degrees of freedom, as these should be included in the results.

### Multiple comparisons
The results so far suggest that molluscicide dosage affects slug activity, but not which treatments differed significantly from each other. To determine which groups significantly differ from each other a posthoc multiple comparisons test, such as the Tukey multiple comparison test, must be used.

To do this the linear model object must first be changed into a different kind of model object using the `aov` function.
```{r}
slug_aov <- aov(slug_model)
```

The Tukey test can be carried out using the `HSD.test` function from the `agricolae` package. Make sure you have loaded the `agricolae` package as above (under First steps) before running the following code. Remember to carry the test out on the `aov` object (not the original fitted model object).

```{r}
HSD.test(slug_aov, "SlugitTreat", console=TRUE)
```

The part of the output that we're really interested in here is the final table. The first and second columns show the treatment names and the mean of the focal variable for that treatment respectively. The third column shows which treatments differed significantly from each other - treatments with the same letter did not differ significantly from each other. For example, in this case the full dose treatment differed significantly from the control and the one-tenth dosage treatment, but not from the half dose treatment (as both the full and half dose treatments have been assigned the letter 'c')

### Presenting the results

The results could be summarised in text as follows:

> There was a significant effect of the dosage of molluscicide on slug activity (ANOVA; F=34.2; d.f.= 3,28; p<0.001) (Figure 1). Slug movement was significantly higher than the control at the one-tenth dose and lower at the half and full doses (Tukey multiple comparison test; p<0.05). Increasing the dosage from a half to a full dose did not have a significant effect on slug activity (Tukey multiple comparison test p>0.05).

Finally make a plot to summarise the results. Start by calculating the mean and standard error of the mean for each group. 
```{r}
slug_stats <- 
  slug %>% 
  group_by(SlugitTreat) %>% 
  summarise(Mean = mean(Distance), SE = sd(Distance)/sqrt(n()))
slug_stats
```

If you want to include the letters from the Tukey test on the plot then add these to the data frame (make sure the order of the letters here corresponds to the order of the treatments in `slug_stats` above.
```{r}
slug_stats <- slug_stats %>% mutate(tukey = c("a", "c", "c", "b"))
```

Change the order that the points will be plotted in so that it increases in dosage.
```{r}
slug_stats <- slug_stats %>% mutate(SlugitTreat = factor(SlugitTreat, levels = c("control", "one tenth dose", "half dose", "full dose")))
```

Then plot the means as points and the standard errors as bars (a bar plot could also be used here using the `geom_bar` function)
```{r, fig.width=4, fig.height =3, fig.align='center'}
ggplot(data = slug_stats, aes(x = SlugitTreat, y = Mean, ymin = Mean - SE, ymax = Mean + SE)) + 
  # this adds the means as points
  geom_point(colour = "blue", size = 3) + 
  # this adds the error bars
  geom_errorbar(width = 0.1, colour = "blue") + 
  # make the labels more informative
  xlab("Molluscicide treatment") + ylab("Distance (cm per min)") +
  # add letters from the Tukey test
  geom_text(aes(x = SlugitTreat, y = Mean + SE + 0.2, label = tukey)) + 
  # flip x and y axes
  coord_flip() +
  # use a more professional theme
  theme_bw()
```

In the figure legend make sure you state what the points, error bars, and letters indicate, for example 

> Figure 1: Effect of molluscicide doseage on the rate of slug movement. Points and error bars show means and standard error of the means respectively. Treatments with the same letter did not differ significantly (Tukey test, p>0.05).
