---
title: Revising APS 135 
author: Dylan Childs
date: Tuesday 3, 2017
highlighter: pygments
output: 
  ioslides_presentation:
    css: presentations.css
    smaller: false
---

```{r, include=FALSE}
showAnswers <- TRUE

library(dplyr)
library(ggplot2)
library(grid)
library(knitr)

storms <- read.csv("../data_csv/STORMS.CSV")

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## While we're waiting {.smaller}

Start RStudio and open up a new R scipt. Add R code to load the **dplyr** and **ggplot2** packages, read in the STORMS.CSV file, and remind yourselfwhat kind of variables it contains, e.g.

```{r, eval = FALSE}
# clear the workspace (aka R's 'brain')
rm(list = ls())

# load and attach the packages we need
library("dplyr")
library("ggplot2")

# read the data in STORMS.CSV into R + call the result 'storms'
storms <- read.csv("STORMS.CSV")
# convert the data frame to a tibble
storms <- as_data_frame(storms)
# take a look at the data
glimpse(storms)
```

The **dplyr** package | Manipulating data 
====================================================================

## The five key verbs {.build .smaller} 

<div>
**dplyr** has five main "verbs", each of which corresponds to a function:

- `select`: Extract a subset of variables
- `filter`: Extract a subset of rows
- `arrange`: Reorder rows 
- `mutate`: Construct new variables 
- `summarise`: Calculate information about groups

(+ a few other important functions that we need too, e.g. `slice`, `rename`, and **most important of all**, `group_by`)
</div>

<div>
<br>

**We can classify the verbs according to what they work on:**

- observations (rows): `filter` & `slice` & `arrange`
- variables (columns): `select` & `rename` & `mutate`

(This classification only works if your data are __tidy__)
</div>

## Using `filter` {.build .smaller}

<div>
Use __relational operators__ to compare the elements of vectors:

- `x < y `: is x less than y?
- `x > y `: is x greater than y?
- `x <= y`: is x less than or equal to y?
- `x >= y`: is x greater than or equal to y?
- `x == y`: is x equal to y?
- `x != y`: is x not equal to y?
</div>

<div>
Use __logical operators__ to combine comparisons:

- `x & y`: are both x AND y TRUE?
- `x | y`: is x OR y true?
</div>

## Using `filter` {.build}

<div>
Use `filter` to select a subset of rows for further analysis based on the result(s) of one or more comparisons. Using `filter` looks like this:

```{r, eval=FALSE}
filter(data, <expression>)
```

__Arguments__

- `data`: the name of a data frame or tibble object  
- `<expression>`: logical comparison involving variables in `data`
</div>

<div>
__Comments__

The `<expression>` typically uses a combination of __relational__ (e.g. `<` and `==`) and (sometimes) __logical__ (e.g. `&` and `|`) operators
</div>

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on more than one variable__
<div id="box-text">
Use the `filter` function on the `storms` data set to select the rows containing just the observations associated with EITHER a 'Hurricane' OR a 'Tropical Storm'. Print the result to the Console.

</div>
</div>

```{r, include=showAnswers}
filter(storms, type == "Hurricane" | type == "Tropical Storm")
```

## Using `select`

We use `select` to to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(data, Variable1, Variable2, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object  
- `VariableX`: the names of one or more variables in `data`

## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function on the `storms` data set to make a new data set containing only `name`, `year` and `pressure`. Assign a name to the result (i.e. use `<-`) and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, include=showAnswers}
simple_storms <- select(storms, name, year, pressure)
glimpse(simple_storms)
```

## Using `mutate` {.build}

<div>
Use `mutate` to __add derived variables__. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(data, NewVar = <expression>, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object
- `NewVar`: name of a new variable to create
- `<expression>`: an expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` that appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`.
- You can use more than one `NewVar = <expression>` at a time if you need to construct several new variables.
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
The sea level pressure of every observation is measured in millibar. Use the `mutate` function on the `storms` data set to make a new variable (inside `storms`) that is the sea level pressure in Pascal. Call the new variable `pressure_Pa`.

Hint: 1 mbar = 100 Pa
</div>
</div>

```{r, include=showAnswers}
mutate(storms, pressure_Pa = 100 * pressure)
```

The **dplyr** package | Piping operations and summarising data 
====================================================================

## Using the 'pipe' `%>%`

The **dplyr** package imports a special operator, called 'the pipe'. It looks like this: `%>%`. This allows us to chain together a sequence of operations, reading from left to right and top to bottom. For example, take a look at this sequence of calculations:

```{r}
x <- 10
x <- sqrt(x)
round(x, 2)
```

<div>
The piped version looks like this

```{r}
10 %>% sqrt() %>% round(2)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
See if you can use the ` %>% ` to construct a pipeline that filters, selects, and then mutates the `storms` data, using the same criteria as were used in each of the last three exercises.
</div>
</div>

```{r, include=showAnswers}
storms %>% 
  filter(type == "Hurricane" | type == "Tropical Storm") %>% 
  select(name, year, pressure) %>% 
  mutate(pressure_Pa = 100 * pressure)
```

## Warning!

Remember that if you want to actually use the result of a piped set of operations, you have to assign a name to the result, e.g.

```{r, eval=FALSE, echo=showAnswers}
processed_storms <- 
  storms %>% 
  filter(type == "Hurricane" | type == "Hurricane") %>% 
  select(name, year, pressure) %>% 
  mutate(pressure_Pa = 100 * pressure)
```

## Using the `summarise` function {.smaller .build}

<div>
We use summarise to __calculate summaries of variables__. This is how we use `summarise`:

```{r, eval=FALSE}
summarise(data, SummaryVar = <expression>, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object
- `SummaryVar`: name of your summary variable
- `<expression>`: an expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` that appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`. However, `<expression>` **must return a single value** (e.g. `mean()`, `var()`, `sd()`, `n()` are all fine).
- Although `summarise` looks a lot like `mutate`, it is designed to construct a completely new, much smaller, data set containing summaries of one or more variables.
- You may use more than one `SummaryVar = <expression>` at a time (separated by `,`) if you need to construct several summaries.
</div>

## Using the `group_by` function {.build .smaller}

<div>
Use `group_by` to add grouping information to a tibble or data frame. This is how we use `group_by`:

```{r, eval=FALSE}
group_by(data, GroupVar1, GroupVar2, ...)
```
</div>

<div>
__Arguments__

- `data`: the name of a data frame or tibble object
- `GroupVar`: name of grouping variable(s) 
</div>

<div>
__Comments__

- The `group_by` function does not do anything more than add grouping information to the input. It is only useful when used with another function (e.g `summarise`).
- Using `group_by` with `summarise` enables us to calculate numerical summaries on a _per group_ basis.

## Exercise {.build}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
You're going to use `group_by` and `summarise` together, along with the pipe (` %>% `) in this exercise. Use the `group_by` and `summarise` functions on the `storms` data set to calculate the minimum and maximum wind speed associated with each month. Call the summary variables `min_speed` and `max_speed`.
</div>
</div>

```{r, include=showAnswers}
storms %>% 
  group_by(month) %>%
  summarise(min_speed = min(wind), max_speed = max(wind))
```

Overview of **ggplot2** | The grammar of graphics
====================================================================

## Key concepts {.build}

You need to wrap your head around a few ideas to start using **ggplot2** effectively:

>- __layers__: We build ggplot2 objects by adding one or more layers together. We can do this in a stepwise way, only plotting the object when we are ready. Each layer needs things like data, aesthetics, geometric objects, etc

>- __aesthetics__: The word aesthetics refers to __the information__ in a plot. For example, the variables associated with: the x and y axes, point colours, line types, etc. We specify this using the `aes` function.

>- __geometric objects__: Geometric objects ("geoms") determine how the information is displayed. For example, will it be a scatter plot or a bar plot? We can specify geoms by adding a layer via functions beginning with `geom_`.

We build up a plot by combining different functions using the `+` operator. This has nothing to do with numeric addition!

## Example: Scatter plots

__Quick reminder of how all this works__:

We don't have to build a plot object up in separate steps and then explicitly "print"" it to the Console. If we just want to make the plot in one go we can do it like this:
```{r, fig.height=2.6, fig.width=3.2}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "purple") + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the last example (a scatter plot of wind speed vs pressure), but now try to customise the point colours, size, and transparency.

Hint: Try `?geom_point` if you can't remember how to do this, looking at the **Aesthetics** section. An example of suitable output is given below.
</div>
</div>

```{r, fig.height=3.1, fig.width=4.0, echo=FALSE}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Answer

```{r, fig.height=3.1, fig.width=4.0, include=showAnswers}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the last example (a scatter plot of wind speed vs pressure), but now make the point colour depend upon storm type (i.e. the `type` variable). This requires that you set up a new mapping between a variable and a suitable 'aesthetic'.
</div>
</div>

```{r, fig.height=3.1, fig.width=5.5, echo=FALSE}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Answer

```{r, fig.height=3.1, fig.width=5.5, include=showAnswers}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise {.build}

<div id="exercise-box">
__Aesthetic mappings vs. arguments to `geom_`__
<div id="box-text">
We can set something like "colour" in two places: the aesthetic mapping (`aes`) or via an argument to a geom (`geom_`). Remember what happens if we set the colour in both places at once? Experiment with the last example to make sure you understand which one---the aesthetic mapping geom argument---has precedence?
</div>
</div>

## Answer

```{r, fig.height=3.1, fig.width=4.0, include=showAnswers}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(colour = "blue", size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## What are histograms and dot plots?

We use histograms and dot plots to understand the __distribution__ of a variable. They summarise the number of observations occurring in a contiguous series of bins. Use `geom_histogram` to construct a histogram and `geom_dotplot`. Here's an example:

```{r, echo = FALSE}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

See if you can make that plot...

## Answer

```{r, fig.height=3.1, fig.width=4.0, include=showAnswers}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

## Using **dplyr** and **ggplot2** together

Dot plots are better than histograms for summarising the distribution of a variable in a small sample. You will see a lot of dot plots this year. See if you can produce one that summarises the distribution of wind speeds associated with storm 'Keith'. To do this, you will need to use **dplyr** to get the right set of observations.

```{r, echo = FALSE}
storms %>%
  filter(name == "Keith") %>%
  ggplot(aes(x = wind)) + 
    geom_dotplot(colour = "darkgrey", fill = "grey", binwidth = 8) + 
    labs(x = "Wind speed", y = "Count") 
```

## Answer

```{r, fig.height=3.1, fig.width=4.0, include=showAnswers}
storms %>%
  filter(name == "Keith") %>%
  ggplot(aes(x = wind)) + 
    geom_dotplot(colour = "darkgrey", fill = "grey", binwidth = 8) + 
    labs(x = "Wind speed", y = "Count") 
```

## Plotting summary statistics (bar plots) {.build}

If we want to compare a summary statistic (e.g. the mean) across groups we need to walk through a two step process:

1. First use functions like `group_by` and `summarise` function from the **dplyr** package to calculate the mean for each group. 

2. Then use **ggplot2** with the new data set of means to plot the summary statistic. We usually use either a bar plot (`geom_col`) or a points plot (`geom_point`) to do this.

## Exercise

<div id="exercise-box">
__Making a barplot of means__
<div id="box-text">
Working with the `storms` dataset, construct a bar plot to summarises the mean wind speed (`wind`) associated with storms in each year (`year`). This is a two step process. First use **dplyr** to calculate the mean wind speed for each year, then use **ggplot2** (`geom_col`) with the summarised data set to produce the bar plot.
</div>
</div>

```{r, fig.width=2.5, fig.height=2, echo=FALSE}
# step 1 - use dplyr to calculate the means
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(wind = mean(wind))
# step 2 - make the plot
ggplot(wind.means, aes(x = year, y = wind)) + 
  geom_col(fill="darkgrey")
```

## Answer

```{r, fig.width=2.5, fig.height=2, include=showAnswers}
# step 1 - use dplyr to calculate the means
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(wind = mean(wind))
# step 2 - make the plot
ggplot(wind.means, aes(x = year, y = wind)) + 
  geom_col(fill="darkgrey")
```

## Exercise (Harder)

<div id="exercise-box">
__Making a barplot of means and standard errors__
<div id="box-text">
Working with the `storms` dataset, construct a bar plot to summarises the mean **and standard error** of wind speed (`wind`) associated with storms in each year (`year`). Again, this is a two step process. First use **dplyr** to calculate the mean wind speed for each year, then use **ggplot2** (`geom_col` and `geom_errorbar`) with the summarised data set to produce the bar plot. N.B. -- you need to add two layers to get both pieces of information onto the figure. 

$$
\text{Standard Error} = \frac{\text{Standard Deviation}}{\sqrt(Sample Size)}
$$
</div>
</div>

```{r, fig.width=2.5, fig.height=2, echo=FALSE}
# step 1
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(mean = mean(wind), 
            se = sd(wind)/sqrt(n()))
# step 2 
ggplot(wind.means, aes(x = year, y = mean, ymin = mean-se, ymax = mean+se)) + 
  geom_col(fill="darkgrey") + 
  geom_errorbar(width = 0.25)
```


## Answer

```{r, fig.width=2.5, fig.height=2, include=showAnswers}
# step 1
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(mean = mean(wind), 
            se = sd(wind)/sqrt(n()))
# step 2 
ggplot(wind.means, 
       aes(x = year, y = mean, ymin = mean-se, ymax = mean+se)) + 
  geom_col(fill="darkgrey") + 
  geom_errorbar(width = 0.25)
```


Wrap up | 
====================================================================

## Miscellaneous 'stuff' {.build}

- R does not change things 'in place'. You have to use assignment ` <- ` ('gets') to store copies of objects...

- Scientific notation in R looks like this: `1e-16`. That is R-speak for $1.0 \times 10^{-16}$, i.e. a very small number.

- Don't forget the basics, e.g. `mean()`, `var()`, `sd()`, `n()`

- Keyboard shortcuts are useful. You should learn a few. I really like 'Alt-Underscore' and 'Ctrl-Shift-M'.

## Two tips for success {.build}

1.  **Keep your scripts organised and well-commented** If you don't, you will make a lot of mistakes, you will find it harder to revise, and next year, you won't have a clue how to analyse your own data.

2.  **Do required reading each week, take good notes, and ask questions** It may be tempting to skip the reading each week. Don't do this. At the very least, work through it quickly, noting anything that is confusing you. 




