---
title: Revising APS 135 
author: Bethan Hindle
date: Tuesday 25th September, 2018
highlighter: pygments
output: 
  ioslides_presentation:
    css: presentations.css
    smaller: false
---

```{r, include=FALSE}
showAnswers <- TRUE

library(dplyr)
library(ggplot2)
library(grid)
library(knitr)

storms <- read.csv("../data_csv/STORMS.CSV")

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## While we're waiting

1. Go to MOLE, download the STORMS.CSV file, and place it somewhere 'sensible' --- i.e. somewhere that you can locate again.

2. Start up RStudio (not R!) and stare at it... Can you remember much from last year? Maybe not... Don't worry, that's what today is all about.

Basics | Setting up and reading in data
===============================================================================

## The working directory {.build .smaller}

A quick reminder of what it is and how to set it...

- The **working directory** is a default location that R uses to looks for files. It is just a folder on your computer.

- If you don't set the working directory, R will do it for you. It probably won't use the folder you want to use.

- **YOU should set the working directory**. Setting your working directory should be the first thing you do EVERY TIME you start R/RStudio.

- Do not do this using R code in your script. Instead, do it using the RStudio menu system (there are other ways, don't use them).

<br><br>

<div>
Set your working directory now using: **Session > Set Working Directory > Choose Directory...**
</div>

## Starting a new script {.build}

Start a new script using: **File > New File... > R Script**

<div>
Then add the following to the beginning of your script:
```{r, eval=FALSE}
# clear the workspace (aka R's 'brain')
rm(list = ls())
```
Can you remember what this does? Why can it be a good idea include it?
</div>

This script hasn't been saved. Do that now...

## Packages {.smaller}

Packages extend what base R can do. Two things have to happen before you can actually use the functions in a package:

1. You have to **install** it. This is a "do once" operation. After a package is installed you can restart R and it is still installed.

2. You have to **load and attach** it. This is a "do every time" operation. You have to load and attach the package every time you start a new session (= open R).

<br><br><br><hr>

A mobile phone App analogy: 

- installing a package <=> getting an App from an App store

- loading and attaching a package <=> opening the App on your phone

## Accessing a package

1. We **install** a package with the `install.packages` function

2. We **load and attach** an installed package with the `library` function

**Remember!**

- Never use `install.packages` inside your scripts! You don't want to re-install a package every time you plan to use it. 

- You can install a package via the Packages tab in RStudio. Use this if you don't like using `install.packages`.

## We need **dplyr** and **ggplot2**...

<div>
**Step 1.** Install **dplyr** and **ggplot2**

Navigate to the Packages tab in RStudio (bottom right pane by default). Can you find **dplyr** and **ggplot2** in the list? 

If not... click the 'Install' button. Type the following: **dplyr**, **ggplot2**... and click 'Install'.
</div>

<div>
**Step 2.** Place the following into your script, and then submit it to the Console:

```{r}
# load and attach the packages we need
library("dplyr")
library("ggplot2")
```
</div>

**Remember: you don't have to repeat 'Step 1' in every new R session**

## Reading a CSV file into R {.build}

This is easy if you've set the working directory properly.

<div>
Include the following line in your script, and submit it to the Console

```{r, eval = FALSE}
# read the data in STORMS.CSV into R + call the result 'storms'
storms <- read.csv("STORMS.CSV")
```
</div>

What happened? If you ran into an error it means the working directory isn't set correctly or the file name is wrong. Get a TA to help you. 

Once it is loaded... Look at the 'Environment' tab in the top right pane. Can you see something called `storms`? What kind of object is `storms`?

Data in R | Vectors, data frames and tibbles...
===============================================================================

## Two views of data frames {.build}

1.   __Data frames are a collection of vectors__. These may be different types (numeric, character, others) but they must all be the same length.

2.   __Data frames are table-like objects__. Take another look at your `storms` data frame. This has rows and columns. 

Data frames are central to data analysis in R. We can think of each vector as corresponding to a __statistical variable__ that we want to analyse. Almost every function we will use this year works with data stored in a data frame.

If every row in the data frame corresponds to a single observation, and every variable is only found in one column, the data are "tidy". We only use tidy data in this course, because this is what R expects most of the time. **Do the same in your own work.**

## Extracting a vector (i.e. a column) {.build}

We can extract a vector from a data frame in a few different ways:

<div>
- Using the name and the `$` operator
```{r, eval = FALSE}
storms$name
```
- Using the name and the `[[ ]]` operator
```{r, eval = FALSE}
storms[["name"]]
```
- Using the position and the `[[ ]]` operator
```{r, eval = FALSE}
storms[[1]]
```
</div>

We'll just use the `$` method, and we won't need it often. 

## Using the `$` operator

Use the `$` operator to extract the `type` variable and print it to the console. 

```{r, eval=showAnswers, echo=showAnswers}
storms$type
```

## 'Tibbles' 

Er, what the @^$£ is a 'tibble'?... Basically, a tibble is a special version of the data frame. The **dplyr** tends to produce them. We can convert a normal data frame to a tibble if we need to, e.g.

```{r, eval=FALSE}
tbl_df(storms)        # <- the old way (deprecated)
# or
as_data_frame(storms) # <- the new way
```

Add this to your script now...

```{r}
# convert the data frame to a tibble
storms <- as_data_frame(storms)
```

The **only** thing you really need to be aware of is that a tibble is printed to the Console more compactly than a data frame.

## Examining data frames {.build .smaller}

We can print the tibble straight to the console...

```{r}
# print the tibble to the console
storms
```

This is quite verbose.

## Examining data frames

Trying using `str` or `glimpse` with `storms`. Then try `View`. Makes sure you can answer the following questions:

 -  How many variables and observations does `storms` contain?
 -  How many different *kinds* of variables does it contain? What are these?

## Examining data frames

Using `str`...

```{r, eval=showAnswers, echo=showAnswers}
str(storms)
```

## Examining data frames

Using `glimpse`...

```{r, eval=showAnswers, echo=showAnswers}
glimpse(storms)
```

## A quick introduction to factors {.build}

By default when we read categorical variables into R it turns them into factors e.g. `name` and `type` in the storms data set.

Let's have a closer look at a factor. Print the `type` variable to your console.

```{r, eval=FALSE, echo=showAnswers}
storms$type
```

Have a look at the last line:

```{r, echo=FALSE}
cat("Levels:", levels(storms$type))
```

What do you think this shows? What do you notice about the ordering?

## A quick introduction to factors {.build}

We can change the order of a factor using `mutate`:

```{r, eval=showAnswers, echo=showAnswers}
storms <- storms %>% 
  mutate(type = factor(type, 
    levels = c("Tropical Depression", "Extratropical", 
               "Tropical Storm", "Hurricane")))
```

Have a look at the last line again

```{r, eval=FALSE, echo=showAnswers}
storms$type
```

```{r, eval=showAnswers, echo=FALSE}
cat("Levels:", levels(storms$type))
```

This can be useful for plotting e.g. here we want to have the storms in order of severity.

## A quick introduction to factors {.build}

Avoid using numbers to encode categorical variables, if you want to avoid making mistakes in R.

I.e. if your treatments are categorical (e.g. feed types) do **not** enter your data like this:

```{r, eval=TRUE, echo=FALSE}
example <- data.frame(Treatment = rep(1:3, each = 3), Count = sample(0:15, 9, replace = TRUE), row.names = NULL)
print(example)

```

## Is your script well-organised?

This is good...

```{r, eval = FALSE}
# clear the workspace (aka R's 'brain')
rm(list = ls())

# load and attach the packages we need
library("dplyr")
library("ggplot2")

# read the data in STORMS.CSV into R + call the result 'storms'
storms <- read.csv("STORMS.CSV")
# convert the data frame to a tibble
storms <- as_data_frame(storms)
# take a look at the data
glimpse(storms)
```

## Is your script well-organised?

This is **not** so good...

```{r, eval = FALSE}
rm(list=ls())
library("dplyr")
library("ggplot2")
storms<-read.csv("STORMS.CSV")
storms<-as_data_frame(storms)
glimpse(storms)
```

## RStudio keyboard shortcuts 

We have seen a few of these already.

- Navigate command history:	Up/Down

- Evaluate an expression: Ctrl+Return

- Insert the assignment operator ` <- `: Alt+Dash(-)

- Insert the pipe ` %>% `: Ctrl+Shift+m

- Move cursor to Source Editor:	Ctrl+1

- Move cursor to Console:	Ctrl+2

Many more here: [https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)


Getting help | Many different routes
===============================================================================

## Built in 'help' files {.smaller}

We do the following:

```{r, eval=FALSE}
?name_of_a_useful_function
```

Help files are really **documentation** files:

- Description
- Usage
- Arguments
- Details
- References
- Examples <-- This is often the most useful part

Better options for help can be found the interweb (e.g. Google, Stack Overflow), books and those cheat sheets... Remember the cheat sheets? Can you find them again?

## Other sources of help {.smaller .build}

The internet is packed full of helpful resources:

- **Google** -- this might be obvious, but you have to work on your "Google-fu" to find what you want. The key is to pick the right search terms:

    - include the function or package name
    
    - include the letter R!
    
    - be selective about which sites you look at...

<div>
- **Stack Overflow** -- probably the best question and answer site for computing stuff. People can be a bit blunt though:

    - check whether your question has been asked before (Google)
    
    - read the community guidelines
    
    - provide example code with your questions
</div>

## Other sources of help {.build}

- **Ask us**

    - Me or the TAs during the workshops
    
    - Attend the drop in help sessions (12-2 on Fridays in B56)
    
    - Send me an email (b.hindle@sheffield.ac.uk)
    
<div>

- **L1 coursebook**

    - http://dzchilds.github.io/eda-for-bio

</div>

## What's next?

- Next week:
  
    - dplyr and ggplot2 revision
  
    - Start reading the coursebook: http://dzchilds.github.io/stats-for-bio/  
    Check which chapters you need to read under Week 2 on MOLE
        
