---
title: Revising APS 135 
author: Dylan Childs
date: Tuesday 27, 2016
highlighter: pygments
output: 
  ioslides_presentation:
    css: presentations.css
    smaller: false
---

```{r, include=FALSE}
showAnswers <- FALSE

library(dplyr)
library(ggplot2)
library(grid)
library(knitr)

storms <- read.csv("../data_csv/STORMS.CSV")

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## While we're waiting

1. Go to MOLE, download the STORMS.CSV file, and place it somewhere 'sensible' --- i.e. a permanent that you can locate again.

2. Start up RStudio (not R!) and stare at it... Can you remember much from last year? Maybe not... Don't worry, that's what today is all about.

Basics | Setting up and reading in data
===============================================================================

## The working directory {.build}

A quick reminder of what it is and how to set it...

- The **working directory** is a default location that R uses to looks for files. It is just a folder on your computer.

- If you don't set the working directory, R will do it for you. It probably won't use the folder you want to use.

- **YOU should set the working directory**. Setting your working directory should be the first thing you do EVERY TIME you start R/RStudio.

- Do not do this using R code in your script. Instead, do it using the RStudio menu system (there are other ways, don't use them).

<br><br>

<div>
Set your working directory now using: **Session > Set Working Directory > Choose Directory...**
</div>

## Starting a new script {.build}

Start a new script using: **File > New File... > R Script**

<div>
Then add the following to the beginning of your script:
```{r, eval=FALSE}
# clear the workspace (aka R's 'brain')
rm(list = ls())
```
Can you remember what this does? Why can it be a good idea include it?
</div>

This script hasn't been saved. Do that now...

## Packages 

Packages extend what base R can do. Two things have to happen before you can actually use the functions in a package:

1. You have to **install** it. This is a "do once" operation. After a package is installed you can restart R and it is still installed.

2. You have to **load and attach** it. This is a "do every time" operation. You have to load and attach the package every time you start a new session (= open R).

<br><br><br><hr>

A mobile phone App analogy: 

- installing a package <=> getting an App from an App store

- loading and attaching a package <=> opening the App on your phone

## Accessing a package

1. We **install** a package with the `install.packages` function

2. We **load and attach** an installed package with the `library` function

**Remember!**

- Never use `install.packages` inside your scripts! You don't want to re-install a package every time you plan to use it. 

- You can install a package via the Packages tab in RStudio. Use this if you don't like using `install.packages`.

## We need **dplyr** and **ggplot2**...

<div>
**Step 1.** Install **dplyr** and **ggplot2**

Navigate to the Packages tab in RStudio (bottom right pane by default). Can you find **dplyr** and **ggplot2** in the list? 

If not... click the 'Install' button. Type the following: **dplyr**, **ggplot2**... and click 'Install'.
</div>

<div>
**Step 2.** Place the following into your script, and then submit it to the Console:

```{r}
library("dplyr")
library("ggplot2")
```
</div>

**Remember: you don't have to repeat 'Step 1' in every new R session**

## Reading a CSV file into R {.build}

This is easy if you've set the working directory properly.

<div>
Include the following line in your script, and submit it to the Console

```{r, eval = FALSE}
storms <- read.csv("STORMS.CSV")
str(storms)
```
</div>

What happened? Look at the 'Environment' tab in the top right pane. Can you see something called `storms`? What kind of object is `storms`?

Data in R | Vectors, data frames, 'tibbles', etc.
===============================================================================

## 'Tibbles' 

Er, what the @^$£ is a 'tibble'?... Basically, a special version of the data frame. **dplyr** tends to produce them. We can convert if we need to, e.g.

```{r, eval=FALSE}
tbl_df(storms)        # <- the old way (deprecated)
# or
as_data_frame(storms) # <- the new way
```

Add this to your script now...

```{r}
storms <- as_data_frame(storms)
```

The **only** thing you really need to be aware of is that a tibble is printed to the Console more compactly than a data frame.

## Examining data frames {.build}

We can print the tibble straight to the console...

```{r}
storms
```

This is quite verbose.

## Examining data frames

Trying using `str` or `glimpse` with `storms`. Then try `View`. Makes sure you can answer the following questions:

 -  How many variables and observations does `storms` contain?
 -  How many different *kinds* of variables does it contain? What are these?

## Examining data frames

Using `str`...

```{r, eval=showAnswers, echo=showAnswers}
str(storms)
```

## Examining data frames

Using `glimpse`...

```{r, eval=showAnswers, echo=showAnswers}
glimpse(storms)
```

## Two views of data frames {.build}

1.   __Data frames are a collection of vectors__. These may be different types (numeric, character, others), but they must all be the same length.

2.   __Data frames are table-like objects__. Take another look at our simple data frame. This clearly has rows and columns. 

Data frames are central to data analysis in R. We can think of each vector as corresponding to a __statistical variable__ that we want to analyse. Almost every function we will learn about this year works with data stored in a data frame.

## Extracting a vector (i.e. a column) {.build}

We can extract a vector from a data frame in a few different ways:

<div>
- Using the name and the `$` operator
```{r, eval = FALSE}
storms$name
```
- Using the name and the `[[ ]]` operator
- Using the position and the `[[ ]]` operator
</div>

We'll just use the `$` method, and we won't need it often. 

## Using the `$` operator

Using the `$` operator to extract the `type` variable and print it to the console. 

```{r, eval=showAnswers, echo=showAnswers}
storms$type
```


Getting help | Many different routes
===============================================================================

## Built in 'help' files

We do the following:

```{r, eval=FALSE}
?name_of_a_useful_function
```

Help files are really **documentation** files:

- Description
- Usage
- Arguments
- Details
- References
- Examples <-- This is often the most useful part

Better options for help are the internet (e.g. Google, stack overflow), books and those cheat sheets...

## Other sources of help {.smaller .build}

The internet is packed full of helpful resources:

1. **Google** -- this might be obvious, but you have to work on your "Google-fu" to find what you want. The key is to pick the right search terms:

    - include the function or package name
    
    - include the letter R!
    
    - be selective about which sites you look at...

<div>
2. **Stack Overflow** -- probably the best question and answer site for computing stuff. People can be a bit blunt though:

    - check whether your question has been asked before (Google)
    
    - read the community guidelines
    
    - provide example code with your questions
</div>


The **dplyr** package | Manipulating data 
====================================================================

## The five key verbs {.build} 

<div>
****dplyr** has five main "verbs", each of which corresponds to a function:**

- `select`: Extract a subset of variables
- `filter`: Extract a subset of rows
- `arrange`: Reorder rows 
- `mutate`: Construct new variables 
- `summarise`: Calculate information about groups

(+ a few other important functions that we need too, e.g. `slice`, `rename`, and **most important of all**, `group_by`)
</div>

<div>
<br>

**It is useful to classify the verbs according to what they work on:**

- observations (rows): `filter` & `slice` & `arrange`
- variables (columns): `select` & `rename` & `mutate`

(This classification only works if your data are __tidy__)
</div>

## Using `filter` {.build .smaller}

**Reminder: comparing values**

<div>
Use __relational operators__ to compare the elements of vectors:

- `x < y `: is x less than y?
- `x > y `: is x greater than y?
- `x <= y`: is x less than or equal to y?
- `x >= y`: is x greater than or equal to y?
- `x == y`: is x equal to y?
- `x != y`: is x not equal to y?
</div>

## Using `filter` {.build}

<div>
Use `filter` to select a subset of rows for further analysis based on the result(s) of one or more comparisons. Using `filter` looks like this:

```{r, eval=FALSE}
filter(data, <expression>)
```

__Arguments__

- `data`: the name of a data frame or tibble object  
- `<expression>`: logical comparison involving variables in `data`
</div>

<div>
__Comments__

The `<expression>` typically uses a combination of __relational__ (e.g. `<` and `==`) operators (+ maybe __logical__ (e.g. `&` and `|`) operators)
</div>

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on more than one variable__
<div id="box-text">
Use the `filter` function on the `storms` data set to select the rows containing just the observations associated with 'Hurricanes' that took place in 1997 or later. Just print the result to the Console.

If you're not sure where to start, use `glimpse` or `str` to remind yourself what the variables in `storms` are called. You need to work with the ones containing information about the storm category and year.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
filter(storms, type == "Hurricane", year >= 1997)
```

## Using `select`

We use `select` to to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(data, Variable1, Variable2, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object  
- `VariableX`: the names of one or more variables in `data`

## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function on the `storms` data set to make a new data set containing only `name`, `year` and `pressure`. Assign a name to the result (i.e. use `<-`) and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
simple_storms <- select(storms, name, year, pressure)
glimpse(simple_storms)
```

## Using `mutate` {.build}

<div>
Use `mutate` to __add derived variables__. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(data, NewVar = <expression>, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object
- `NewVar`: name of a new variable to create
- `<expression>`: an expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` that appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`.
- You can use more than one `NewVar = <expression>` at a time (separated by `,`) if you need to construct several new variables.
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
The sea level pressure of every observation is measured in millibar. Use the `mutate` function on the `storms` data set to make a new variable (inside `storms`) that is the sea level pressure in Pascal. Call the new variable `pressure_Pa`.

Hint: 1 mbar = 100 Pa
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
mutate(storms, pressure_Pa = 100 * pressure)
```

The **dplyr** package | Piping operations and summarising data 
====================================================================

## Using the 'pipe' `%>%`

The **dplyr** package imports a special operator, called 'the pipe'. It looks like this: `%>%`. This allows us to chain together a sequence of operations, reading from left to right and top to bottom. For example, take a look at this sequence of calculations:

```{r}
x <- 10
x <- sqrt(x)
round(x, 2)
```

<div>
The piped version looks like this

```{r}
10 %>% sqrt() %>% round(2)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
See if you can use the ` %>% ` to construct a pipeline that filters, selects, and then mutates the `storms` data, using the same criteria as were used in each of the last three exercises.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
storms %>% 
  filter(type == "Hurricane", year >= 1997) %>% 
  select(name, year, pressure) %>% 
  mutate(pressure_Pa = 100 * pressure)
```

## Warning!

Remember that if you want to actually use the result of a piped set of operations, you have to assign a name to the result, e.g.

```{r, eval=FALSE, echo=showAnswers}
processed_storms <- 
  storms %>% 
  filter(type == "Hurricane", year >= 1997) %>% 
  select(name, year, pressure) %>% 
  mutate(pressure_Pa = 100 * pressure)
```

## Using `summarise` {.build}

<div>
We use summarise to __calculate summaries of variables__. This is how we use `summarise`:

```{r, eval=FALSE}
summarise(data, SummaryVar = <expression>, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object
- `SummaryVar`: name of your summary variable
- `<expression>`: an expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` that appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`. However, `<expression>` **must return a single value** (e.g. `mean()`, `var()`, `sd()`, `n()` are all fine).
- Although `summarise` looks a lot like `mutate`, it is designed to construct a completely new, much smaller, data set containing summaries of one or more variables.
- You may use more than one `SummaryStat = <expression>` at a time (separated by `,`) if you need to construct several summaries.
</div>

## Using `group_by` {.build}

<div>
Use `group_by` to add grouping information to a tibble or data frame. This is how we use `group_by`:

```{r, eval=FALSE}
group_by(data, GroupVar1, GroupVar2, ...)
```
</div>

<div>
__Arguments__

- `data`: the name of a data frame or tibble object
- `GroupVar`: name of grouping variable(s) 
</div>

<div>
__Comments__

- The `group_by` function does not do anything more than add grouping information to the input. It is only useful when used with another function (e.g `summarise`).
- Using `group_by` with `summarise` enables us to calculate numerical summaries on a _per group_ basis.

## Exercise {.build}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
You're going to use `group_by` and `summarise` together, along with the pipe (` %>% `) in this exercise. Use the `group_by` and `summarise` functions on the `storms` data set to calculate the minimum and maximum wind speed associated with each month. Call the summary variables `min_speed` and `max_speed`.

Hint: You need to know the names of the two functions that calculate the minimum and maximum of a numeric vector.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
storms %>% 
  group_by(month) %>%
  summarise(min_speed = min(wind), max_speed = max(wind))
```

Overview of **ggplot2** | The grammar of graphics
====================================================================

## Key concepts {.build}

You need to wrap your head around a few ideas to start using **ggplot2** effectively:

>- __layers__: We build ggplot2 objects by adding one or more layers together. We can do this in a stepwise way, only plotting the object when we are ready. Each layer needs things like data, aesthetics, geometric objects, etc

>- __aesthetics__: The word aesthetics refers to __the information__ in a plot. For example, the variables associated with: the x and y axes, point colours, line types, etc. We specify this using the `aes` function.

>- __geometric objects__: Geometric objects ("geoms") determine how the information is displayed. For example, will it be a scatter plot or a bar plot? We can specify geoms by adding a layer via functions beginning with `geom_`.

We build up a plot by combining different functions using the `+` operator. This has nothing to do with numeric addition!

## Example: Scatter plots

__Quick reminder of how all this works__:

We don't have to build a plot object up in separate steps and then explicitly "print"" it to the Console. If we just want to make the plot in one go we can do it like this:
```{r, fig.height=2.6, fig.width=3.2}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "purple") + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the last example (a scatter plot of wind speed vs pressure), but now try to customise the point colours, size, and transparency.

Hint: Try `?geom_point` if you can't remember how to do this, looking at the **Aesthetics** section. An example of suitable output is given below.
</div>
</div>

```{r, fig.height=3.1, fig.width=4.0, echo=FALSE}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Answer

```{r, fig.height=3.1, fig.width=4.0, eval=showAnswers, echo=showAnswers}
ggplot(storms, aes(x = pressure, y = wind)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the last example (a scatter plot of wind speed vs pressure), but now make the point colour depend upon storm type (i.e. the `type` variable). This requires that you set up a new mapping between a variable and a suitable 'aesthetic'.
</div>
</div>

```{r, fig.height=3.1, fig.width=5.5, echo=FALSE}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Answer

```{r, fig.height=3.1, fig.width=5.5, eval=showAnswers, echo=showAnswers}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## Exercise {.build}

<div id="exercise-box">
__Aesthetic mappings vs. arguments to `geom_`__
<div id="box-text">
We can set something like "colour" in two places: the aesthetic mapping (`aes`) or via an argument to a geom (`geom_`). Remember what happens if we set the colour in both places at once? Experiment with the last example to make sure you understand which one---the aesthetic mapping geom argument---has precedence?
</div>
</div>

## Answer

The `geom_` has precedence...

```{r, fig.height=3.1, fig.width=4.0, eval=showAnswers, echo=showAnswers}
ggplot(storms, aes(x = pressure, y = wind, colour = type)) + 
  geom_point(colour = "blue", size = 2, alpha = 0.3) + 
  labs(x = "Pressure (mbar)", y = "Wind speed (km/hr)")
```

## What are histograms and dot plots?

We use histograms and dot plots to understand the __distribution__ of a variable. They summarise the number of observations occurring in a contiguous series of bins. Use `geom_histogram` to construct a histogram and `geom_dotplot`. Here's an example:

```{r, echo = FALSE}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

See if you can make that plot...

## Answer

```{r, fig.height=3.1, fig.width=4.0, eval=showAnswers, echo=showAnswers}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

## Using **dplyr** and **ggplot2** together

Dot plots are better than histograms for summarising the distribution of a variable in a small sample. You will see a lot of dot plots this year. See if you can produce one that summarises the distribution of wind speeds associated with storm 'Keith'. To do this, you will need to use **dplyr** to get the right set of observations.

```{r, echo = FALSE}
storms %>%
  filter(name == "Keith") %>%
  ggplot(aes(x = wind)) + 
    geom_dotplot(colour = "darkgrey", fill = "grey", binwidth = 8) + 
    labs(x = "Wind speed", y = "Count") 
```

## Answer

```{r, fig.height=3.1, fig.width=4.0, eval=showAnswers, echo=showAnswers}
storms %>%
  filter(name == "Keith") %>%
  ggplot(aes(x = wind)) + 
    geom_dotplot(colour = "darkgrey", fill = "grey", binwidth = 8) + 
    labs(x = "Wind speed", y = "Count") 
```

## Plotting summary statistics (bar plots)






Wrap up | 
====================================================================

## Miscellaneous 'stuff' {.build}

- R does not change things 'in place'. You have to use assignment ` <- ` ('gets')...

- Scientific notation in R looks like this: `1e-16`. That is R-speak for $1.0 \ times 10^{-16}$, i.e. a very small number.

- Don't forget the basics, e.g. `mean()`, `var()`, `sd()`, `n()`

- Keyboard shortcuts are useful. We really like 'Alt-Underscore' (` <- `) and 'Ctrl-Shift-M' (` %>% `).

## Two tips for success {.build}

1.  **Keep your scripts organised and well-commented** If you don't, you will make more mistakes, you will find it harder to revise, and next year, you won't have a clue how to analyse your own data.

2.  **Read the required chapters each week, take good notes, and ask questions** People who do all three of these things tend not to struggle in this course.




