```{r, eval=showAnswers, echo=showAnswers}
filter(storms, type == "Hurricane", year >= 1997)
```

Using `select` | Extracting a subset of variables 
====================================================================

## Basic usage

We use `select` to to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(data, Variable1, Variable2, ...)
```

__Arguments__

- `data`: the name of a data frame or tibble object  
- `VariableX`: the names of one or more variables in `data`

## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function on the `storms` dataset to make a new data set containing only `name` and `year`. Assign a name to the result (i.e. use `<-`) and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
simple_storms <- select(storms, name, year)
glimpse(simple_storms)
```

Using `mutate` | Making new variables 
====================================================================

## Basic usage {.build}

<div>
Use `mutate` to __add derived variables__. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(data, NewVar = <expression>, ...)
```
</div>

<div>
__Arguments__

- `data`: the name of a data frame or tibble object
- `NewVar`: name of a new variable to create
- `<expression>`: an expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` that appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`.
- You can use more than one `NewVar = <expression>` at a time (seperated by `,`) if you need to construct several new variables.
</div>







## Remember

-   You have to use assignment ` <- ` ('gets')...

