### How big does a bootstrap sample need to be?

Use the R code above to investigate how large a bootstrap sample needs to be to reliably estimate the standard error. Start with a bootstrap sample size of 10, and gradually increase it, calculating the standard error each time you do so. How many bootstrap samples are needed to reliably estimate the SE to one decimal place?

## Week 4: *t*-tests

You should work through the exercises step-by-step, following the instructions carefully. At various points we will interupt the flow of instructions with a question. Make a note of your answer so that you can complete the associated MOLE quiz, which is called 't-tests 1'.   

### Eagle owls and Norway rats

```{r, eval = FALSE, echo = FALSE}
ratskulls <- read.csv("./data_csv/RATSKULL.CSV")
ggplot(ratskulls, aes(x = Weight)) + 
  geom_dotplot() + facet_wrap(~Season, ncol = 1)
t.test(Weight ~ Season, data = ratskulls)
```

A data set containing information about the sizes of Norway rat skulls in the pellets of Scandinavian eagle-owls is available in the RATSKULL.CSV file (you may have come across this before). The data comprise a column of rat skull sizes (measured in grams) and a column of codes indicating the season when a particular skull sample was taken. These data were collected in order to evaluate whether there is a difference between sizes of rats eaten in summer and winter. That is, we want to know if there is a statistically significant difference between the mean rat skull sizes in the winter and summer samples.

Download the RATSKULL.CSV file from MOLE and **place it in your working directory** (this is the location you set at the beginning of this practical). Read the data in RATSKULL.CSV into an R data frame, remembering to assign the data frame a name. Ask a demonstrator to remind you how to do this if you have forgotten, or look over the instructions for reading in the MORPH_WEIGHTS.CSV file from the last practical.

As always, we should always start by *looking at the data* — both visually and in terms of its descriptive statistics:

**Inspection.** Use the `View` function and `dplyr` function `glimpse` to visually inspect the raw data. What are the names given to rat skull size variable and the season indicator variable? What values does the season indicator variable take?

**Descriptive statistics.** Use the appropriate `dplyr` functions (`group_by` and `summarise`) to calculate the sample size, sample mean and standard deviation of each sample. HINT: you will need to use the `mean`, `sd` and `length` functions to help you do this.

**Graphs.** Use `ggplot2` to construct a pair of dot plots, one above the other, to summarise the winter and summer skull size distributions. HINT: you will need to use `geom_dotplot` and the `facet_wrap` functions to do this; look over the plant morph example from the beginning of this practical to see how to use these.

Using the dot plots, and the descriptive statistics, conduct an informal evaluation of the assumptions of the t-test. You should re-read the relevant section above if you can't remember what these are.

```{block, type='do-something'}
**MOLE question**

Do you feel the data conform acceptably to the assumptions? If not, explain why.
```

Let's carry on, assuming that we are confident that it is OK to use a two sample t-test to compare the sample means. Use the R `t.test` function to carry out this evaluation now.

```{block, type='do-something'}
**MOLE question**

Write a concise but complete conclusion summarizing the results of the test.

Is this what you expected from looking at the distributions of data in the two samples?

Suggest two possible biological reasons for the result you observe.
```

### Fungal infection in French beans

```{r, echo=FALSE, eval=FALSE}
sandy <- c(2.3, 2.4, 2.5, 2.6, 2.8, 2.7, 3.1, 2.3, 2.5)
clay  <- c(2.3, 2.5, 2.8, 3.2, 2.9, 3.1, 3.2)
t.test(sandy, clay)
beans <- data.frame(glucosamine = c(sandy, clay),
                    soil = rep(c("sandy","clay"), c(length(sandy), length(clay))))
write.csv(beans, file = "./course-data/FRENCH_BEANS.CSV", row.names = FALSE)
```

A plant pathologist noticed that fungal infection in roots of French beans (*Phaseolus vulgaris*) was rather variable among crops and hypothesized that infection might be affected by the soil type: in particular whether the beans were grown on clayey or sandy soils. Root samples were taken from beans growing in each soil type and fungal infection was measured indirectly by measuring the amount of glucosamine in the roots. Glucosamine is a fungal sugar which is polymerised into chitin which forms the cell walls of most fungi.

The glucosamine concentrations ($\mu$g g$^{-1}$ root dry weight) recorded from the samples were:

  ------------ ----- ----- ----- ----- ----- ----- ----- ----- -----
  Sandy soil    2.3   2.4   2.5   2.6   2.8   2.7   3.1   2.3   2.5
  Clay soil     2.3   2.5   2.8   3.2   2.9   3.1   3.2        
  ------------ ----- ----- ----- ----- ----- ----- ----- ----- -----

Download the FRENCH_BEANS.CSV file from MOLE and place it in your working directory. Read the data into an R data frame (remember to give this a name!), inspect the data, generate some summary statistics (means and SDs) and then plot the data, just as you did in the last exercise. This should be quick to achieve--just copy and paste the code you produced, and edit this where required.

Use a t-test to decide whether there is a significant difference between the amount of infection of bean roots in the two different soils.

```{block, type='do-something'}
**MOLE question**

Make a note of the results:

Mean for plants on clay soil = ?

Mean for plants on sandy soil = ?

t = ?

d.f. = ? 

p = ? 
```

```{block, type='do-something'}
**MOLE question**

Write a statement of the result of the test suitable for inclusion in the results section of the plant pathologist's report.
```

### The power of pairing

The paired *t*-test is a very powerful technique. You can get an idea of the value of the paired-sample test by seeing what would have happened if we had ignored the pairing structure of data and analysed it with an unpaired, two-sample approach (N.B.---This analysis is wrong!):
```{r}
t.test(Glycolipid ~ Drug, data = glycolipid)
```

<div class="well">
**MOLE question**

What result do you get and how does this compare with the paired-sample test?
</div>

### Sheep, grass and nature reserves

The management committee of a nature reserve wants to manage some large grassland areas of the reserve using low density sheep grazing to prevent the grass becoming too long and making the habitat unsuitable for some of the low-growing herbaceous plants for which the reserve is important. Before implementing the plan they conduct a pilot experiment using some fenced plots on the reserve, to test whether low density sheep grazing affects various species of plants.

One problem is that the area is very variable - some parts are wetter than others, and the plants of interest are not particularly evenly distributed. There is also a limit to the number of plots (and sheep) they can use in the experiment. In order to make the maximum use of the resources and, take some account of the variability in the habitat the experiment is set up by randomly placing eight fenced plots around the reserve, with each plot being divided in half by a fence down the middle. Sheep are introduced to one half of each plot (the half being randomly selected in each case), and allowed to graze for the appropriate period of the year. The other half is left ungrazed.

```{block, type='do-something'}
**MOLE question**

Why is this a better design than just having separate grazed and ungrazed plots positioned at random?
```

Over the next 2 years, the abundances of various plants in the in the plots are surveyed.

The data below give the numbers of gentians from each of the eight half-plots with sheep, and the corresponding ungrazed halves after one year of the experiment. 

  ----------- -------- -------- -------- -------- -------- -------- -------- --------
  Treatment                                                                  
               Plot 1   Plot 2   Plot 3   Plot 4   Plot 5   Plot 6   Plot 7   Plot 8
  Grazed         27       1        16       8        10       19       30       9
  Ungrazed       14       6        17       5        0        11       21       6
  ----------- -------- -------- -------- -------- -------- -------- -------- --------
  
These data are stored in GENTIANS_GRAZING.CSV. Test whether there is any evidence for an effect of sheep grazing on the numbers of gentians.

```{block, type='do-something'}
**MOLE question**

What is your conclusion?
```

```{block, type='do-something'}
**MOLE question**

What other comparison would it be useful to be able to make in order to reach a satisfactory conclusion about the effects of grazing?  What test would you do for this?
```

## Regression

### Partridges and hedgrows {#exercise}

```{block, type='do-something'}
Make a note of your answers so that you can complete the associated MOLE quiz, which is called 'regression 1'.   
```

```{r, echo = FALSE, eval = FALSE}
partridge <-read.csv(file = "./course-data/PARTRIDG.CSV")
ggplot(partridge, aes(x = Hedgerow, y = Partridge)) + geom_point()
partridge.model <- lm(Partridge ~ Hedgerow, data = partridge)
anova(partridge.model)
summary(partridge.model)
plot(partridge.model)
```

Hedgerows are the main nesting habitat of the grey partridge (*Perdix perdix*). A survey was carried out to establish whether the abundance of hedgerows in agricultural land had an effect on the abundance of grey partridge. From an area of agricultural land covering several farms, twelve plots were selected which had land uses as similar as possible but differed, as evident from preliminary inspection, in the density of hedgerows (km hedgerow per km^2^). Plots were deliberately selected to cover a wide range of hedgerow densities. The total hedgerow lengths, and exact plot areas, were measured by use of large scale maps. The density of partridges was established by visiting all fields in a study plot once immediately after dawn and once just before dusk, when partridges are feeding and therefore most likely to be seen. Counts of birds observed were made on each visit and the dawn and dusk data were averaged to give a value for partridge abundance for each study plot.

The data are stored in a CSV file PARTRIDG.CSV. The density of hedgerows (km per km^2^) is in the `Hedgerow` variable and the density of partridges (no. per km) is in the `Partridge` variable. Read in the data and take a look at it using the `View` function.

```{block, type='do-something'}
**MOLE question**

Which way round should the variables be?

-   Independent ($x$):

-   Dependent ($y$):
```

Make a scatter plot to allow you to evaluate the assumptions.

```{block, type='do-something'}
**MOLE question**

1.  If there is a relationship does it look linear? 

2.  Ratio or interval data? 

3.  Independent variable $y$ likely to be normally distributed for each $x$?

4.  Variance increases or decreases markedly with increasing $x$?

5.  Errors in $x$ likely to be small compared to those in $y$?
```

If everything is OK, or roughly so, then carry out a regression on the relationship between hedgerow density and partridge density.

```{block, type='do-something'}
**MOLE question**

Summarise the results of your analysis in words.
```

Finish up by preparing a figure that summarises the data and the best fit line estimated from the regression.


